results

### 2020-07-18 analysis/results.Rmd
development of elseviers oa journal portfolio

---
title: "results"
author: "Najko Jahn"
date: "7/7/2020"
output: github_document
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center",
  dpi = 300
)
options(scipen = 999, digits = 4)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

@@

to do

#### What is the uptake of open access in Elsevier's hybrid journal portfolio?

@@

```{r}
library(tidyverse)
library(here)
els_jns_df <- readr::read_csv(here::here("data", "elsevier_apc_list.csv"))
```

To determine the number of Elsevier hybrid journals with at least one open access article, we downloaded and parsed the publisher's APC pricing list from 31 May 2020. The list comprised `r length(unique(els_jns_df$issn))` journals, of which `r els_jns_df %>% filter(oa_type == "Hybrid") %>% distinct(issn) %>% nrow()` supported the hybrid model, representing a share of `r round(els_jns_df %>% filter(oa_type == "Hybrid") %>% distinct(issn) %>% nrow() / length(unique(els_jns_df$issn)) * 100, 0)`%. 
```{r}
hybrid_md <- readr::read_csv(here::here("data", "hybrid_md.csv"))
els_tdm <- readr::read_csv(here::here("data", "els_tmd_links.csv"))
invoice_df <-  readr::read_csv(here::here("data", "invoice_df.csv"))

@@

Of those, `r hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow()` hybrid journals published at least one open access article immediately under a Creative Commons license between 2015 and 2019, corresponding to about `r round(hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow() / els_jns_df %>% filter(oa_type == "Hybrid") %>% nrow() * 100, 0)`% of journal titles in Elsevier’s hybrid journal portfolio. At the same time, `r 100 - round(hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow() / els_jns_df %>% filter(oa_type == "Hybrid") %>% nrow() * 100, 0)`% of hybrid journals did not published an open access article in this period and were, thus, excluded from further analysis.

>>>>

---
title: "results"
author: "Najko Jahn"
date: "7/7/2020"
bibliography: references.bib
csl: chicago.csl
output: github_document
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center",
  dpi = 300
)
options(scipen = 999, digits = 4)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

```{r}
# libraries
library(tidyverse)
library(here)
library(cowplot)
```

@@

Using text-mined data from web-archived Elsevier APC pricing lists  [@Matthias_2020], we were able to determine the evolution of Elsevier's open access journal portfolio including hybrid journals. Figure shows that the majority of journals supported the hybrid model. The year 2016 marked the highest number of journals. After that, the trend reversed. Since 2018, the number of journals stagnated.

```{r}
els_jns <- readr::read_csv(here::here("data", "els_historic_jns.csv")) 
els_jns_count <- els_jns %>%
  group_by(year, oa_model) %>%
  count(year) %>%
  group_by(year) %>%
  mutate(prop = n / sum(n)) %>%
  mutate(oa_model_fct = fct_rev(as_factor(oa_model))) 
ggplot(els_jns_count,
       aes(as.character(year), n,  fill = oa_model_fct, group = oa_model_fct)) +
  geom_area(alpha = 0.8,
            colour = "white") +
  scale_fill_manual(
    values = c("#cccccca0", "#56b4e9"),
    name = NULL,
    labels = c("Fully", "Hybrid")
  ) +
  labs(x = "Year", y = "Elsevier Journals") +
  scale_y_continuous(
    labels = scales::number_format(big.mark = ","),
    expand = expansion(mult = c(0, 0.05)),
    breaks =  scales::extended_breaks()
  ) +
  theme_minimal_hgrid() +
  theme(legend.position = "top",
        legend.justification = "right") +
  guides(fill = guide_legend("OA Business Model"))
```

Compared to 2015, the number of hybrid journals grew by  `r (els_jns_count %>% filter(oa_model == "Hybrid", year == 2020) %>% .$n) - (els_jns_count %>% filter(oa_model == "Hybrid", year == 2015) %>% .$n)` journals, while its proportion relative to Elsevier's overall journal portfolio slightly decreased from `r round(els_jns_count %>% filter(oa_model == "Hybrid", year == 2015) %>% .$prop, 2) * 100`% to `r round(els_jns_count %>% filter(oa_model == "Hybrid", year == 2020) %>% .$prop, 2) * 100`%.

```{r}
flipped_jns <- els_jns %>%
  distinct(issn, oa_model) %>%
  group_by(issn) %>%
  filter(n() > 1)
flipped_jns_ind <- els_jns %>%
  filter(issn %in% flipped_jns$issn) %>%
  group_by(issn) %>%
  slice(which.max(year)) %>%
  ungroup() %>%
  count(oa_model)
```

Following the approach of an earlier study [@Matthias_2019], we furthermore used APC pricing list information to determine the number of open access business model flips between 2015-2020. Although the transition from hybrid to a fully open access business model has drawn most attention, also the reverse can be observed. A total number of `r sum(flipped_jns_ind$n)` journals changed the business model under the ownership of Elsevier. Of these, `r filter(flipped_jns_ind, oa_model == "Open Access")$n` journals flipped to fully open access, while `r filter(flipped_jns_ind, oa_model != "Open Access")$n`changed to hybrid open access. Examples include the physical journal *Materials Today* and the life science journal *Meta Gene*.

#### What is the uptake of open access in Elsevier's current hybrid journal portfolio?

@@

<!-- here, we change to the most current pricing list, which seems to be identical to 31 May-->

```{r}
els_jns_df <- readr::read_csv(here::here("data", "elsevier_apc_list.csv"))
```

```{r}
hybrid_md <- readr::read_csv(here::here("data", "hybrid_md.csv"))
els_tdm <- readr::read_csv(here::here("data", "els_tmd_links.csv"))
invoice_df <-  readr::read_csv(here::here("data", "invoice_df.csv"))

@@

To determine the number of Elsevier hybrid journals with at least one open access article, we downloaded and parsed the publisher's APC pricing list from 31 May 2020. The list comprised `r length(unique(els_jns_df$issn))` journals, of which `r els_jns_df %>% filter(oa_type == "Hybrid") %>% distinct(issn) %>% nrow()` supported the hybrid model, representing a share of `r round(els_jns_df %>% filter(oa_type == "Hybrid") %>% distinct(issn) %>% nrow() / length(unique(els_jns_df$issn)) * 100, 0)`%. Of those, `r hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow()` hybrid journals published at least one open access article immediately under a Creative Commons license between 2015 and 2019, corresponding to about `r round(hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow() / els_jns_df %>% filter(oa_type == "Hybrid") %>% nrow() * 100, 0)`% of journal titles in Elsevier’s hybrid journal portfolio. At the same time, `r 100 - round(hybrid_df %>% filter(oa_archive == FALSE) %>% distinct(issn) %>% nrow() / els_jns_df %>% filter(oa_type == "Hybrid") %>% nrow() * 100, 0)`% of hybrid journals did not published an open access article.
